Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2022-01-27 analysis/paper/paper.Rmd
updated some text

While the overall structure of the model seems to group colourless and white materials towards the bottom left, with dark material to the right, there are a number of samples that break this pattern. This suggests that the perceived colour of a material does not necessarily match what is detected within the visible spectrum of light. It is also difficult to separate out the greyscale of colours that occur inbetween the "black and white" groups, where both light gray and semi-translucent samples end up. All in all, the results echo the complexity and unreliability that exists within the topic of petrographic colour.

@@

## NIR - near infrared range
In the case of the 1 000 nm - 2 500 nm band of the NIR data, similar features can be observed in the average spectra for all four colour groups, with darker materials exhibiting higher absorbance and the light group the most distinct peaks. The presence of molecular water can be inferred based on the peaks at 1 400 nm and 1 900 nm  [@hunt77; @sciuto_etal19] (fig. \@ref(fig:nir-spectra)). Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter falling within one of the stronger CO^3^ bands [@clark_etal90]. Inspection of individual spectra for the light group, however, also reveals a 5th peak for some samples at 2 450 nm. Comparisons to spectra collected on muscovite [@clark_etal90 p.12 675] shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm. It is therefore possible that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other samples Al-OH, possibly inidicating the presence of muscovite.

@@

## Raman
There are two distinctive features present in the Raman data (fig. \@ref(fig:raman-spectra)). The first are three peaks in the 100 cm^-1^ - 500 cm^-1^ band that are typical for SiO~2~ and quartz-based materials, with the strongest peak occuring at 460 cm^-1^ [@gillet_etal90]. These peaks are present for all samples in the dataset. The second feature shows up as two peaks occurring at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite [@cesare_maineri99]. The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ~ 695 cm^-1^ and 795 cm^-1^. Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks are present in measurements collected on, among others, albite, apatite, diabase and shale. Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm^-1^ band. Considering the petrogenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.

@@

```{r raman-prep-data}
#aggregate observations by group(sample) and calculate average of wavelength measurements
raman.averaged <- 
  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)

>>>>

While the overall structure of the model seems to group colourless and white materials towards the bottom left, with dark material to the right, there are a number of samples that break this pattern. This suggests that the perceived colour of a material does not necessarily match what is detected within the visible spectrum of light. It is also difficult to separate out the greyscale of colours that occur inbetween the "black" and "white" groups, where both light gray and semi-translucent samples end up. All in all, the results echo the complexity and unreliability that exists within the topic of petrographic colour.

@@

## NIR - near infrared range
In the case of the 1 000 nm - 2 500 nm band of the NIR data, similar features can be observed in the average spectra for all four colour groups, with darker materials exhibiting higher absorbance and the light group the most distinct peaks. The presence of molecular water can be inferred based on the peaks at 1 400 nm and 1 900 nm  [@hunt77; @sciuto_etal19] (fig. \@ref(fig:nir-spectra)). Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter falling within one of the stronger CO^3^ bands [@clark_etal90; @ramanaidou15]. Inspection of individual spectra for the light group, however, also reveals a 5th peak for some samples at 2 450 nm. Comparisons to spectra collected on muscovite [@clark_etal90 p.12; @ramanaidou15 p.220] shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm. It is therefore possible that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other samples Al-OH, possibly inidicating the presence of muscovite.

@@

## Raman
There are two distinguishing features present in the Raman data (fig. \@ref(fig:raman-spectra)). The first are three peaks in the 100 cm^-1^ - 500 cm^-1^ band that are typical for SiO~2~ and quartz-based materials, with the strongest peak occuring at 460 cm^-1^ [@gillet_etal90]. These peaks are present for all samples in the dataset. The second feature shows up as two peaks occurring at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite [@cesare_maineri99]. The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ~ 695 cm^-1^ and 795 cm^-1^. Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks are present in measurements collected on, among others, albite, apatite, diabase and shale. Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm^-1^ band. Considering the petrogenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset, although measurements on iron ore minerals have also generated a response in the 600 - 700 cm^-1^ band region [@ramanaidou15].

@@

```{r raman-prep-data, results="hide", fig.keep="none"}
#aggregate observations by group(sample) and calculate average of wavelength measurements
raman.averaged <- 
  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
